Katana: A Hot Patching Framework for ELF Executables

### 2010-03-17 paper1/hotpatch/patching.tex
updated paper with more info about safety, limitations of type patching, thoughts on multithreading

For dynamic symbols, Katana traverses the PLT entries of the
executable and compares the symbol name of each entry with
the symbol name that requires relocation. Once a match is found, the
symbol value can be determined. In our prototype,
Katana is unable to handle calls to previously
%SB:
unused functions present in any shared libraries\footnote{This would
  require creation of new PLT and GOT entries and either subsequent
  re-basing of the following segments of the executable, or creation
  of a new segment to allocate the extra entries. Although ELF
  rewriting systems like ERESI or Diablo show that such manipulations
  can be made practical, we chose not to complicate Katana with
  them yet.}
, although support for this is planned.

@@

(i) {\it Type Discovery:} We set out to discover all newly created or
modified data types -- those that are primarily user-defined (such as
structures and unions in {\bf C}). Katana traverses the type information
(as identified by the above DWARF tags) from the newly created executable,
 and for each encountered type, it searches for
the corresponding type-name within the {\it replicated} executable
(from before the patch). If so found, the full types (i.e. the number,
type and position of all member variables contained within) are
compared to determine if they are identical. If not identical,
a transformation between the old and new versions is generated, along
with DWARF information identifying the type to insert into the PO as
soon as a variable is found making use of the altered type.
%the parent type identifying the struct/union is inserted into the PO.
 Else, if
the type name itself was not found within the replicated executable, then the
current type was created by the patch, and is added as such to the PO.
%\newline
\newline
(ii) {\it Data Traversal:} The next step is to traverse all variables
defined within the new application, and for each one encountered,
we first determine its 
%SB:
lexical scope.
If the scope is
local, then we ensure that the corresponding function (the one that
defines this variable) does not have an activation frame on  
the program stack while
applying the patch. Else, the variable has been defined as either
global or static. We first check if the replicated executable defines
the same variable. If not, then this variable has been created by the
patch and we need not worry about it and leave the symbol resolution upto
the compiler (as only {\it new} code can use this variable). Otherwise,
we verify whether the variable's type is one of the
modified types identified during {\it type discovery}. If it is, then
we add the variable along with its {\it original} address from the replicated
executable, its {\it new} address from the patch, and type information to the
PO. At the end of this stage, Katana would have identified all newly
created or modified variables from the patch. Dealing with variable
scope and safety is still a work in progress in the current prototype.
%\newline
\newline
(iii) {\it Patch Application: } Applying a data patch consists of
first tracking down the relevant symbols in program memory. Katana
reads in the PO, and for each data variable encountered, it checks if
the variable is a pointer or not. If it is, then the current validity of the
pointer is verified (by bounds-checking the pointer value to within
heap boundaries).  If the pointer is found to be invalid, no further
action is taken.  If the pointer is valid, then memory for the new
type(s) is allocated, the older structure is copied into the new one
taking into account the difference in structure definition, the old
memory is then freed, and the pointer is modified to point to the new
segment (in case of structures such as lists, trees, since we have the
type specification, we can repeat this process recursively for each
node on the list or tree). Else if the variable is not a pointer, then
Katana modifies all its references in the program text to the updated
memory location from the patch. In the current prototype, Katana
automatically zeros all {\it new} member variables within structures.
Eventually it will support default values and programmer-written
custom initializers.

{\bf Challenges.} %\hskip 0.1in 
Hot patching still faces a number of
challenges, including dealing with multithreaded programs and address
space randomization (which slight changes to the OS loader can help us
overcome).  
There is nothing inherent in our design which will not work with
multithreading, but dealing with it through ptrace takes considerable
work which we have not yet done.
Furthermore, dynamic updates require some knowledge of the
program's execution state so that the application is quiescent with
respect to the code and data being altered by Katana as it applies the
Patch Object. 
% We consider the program to be in a safe state if all
% activation records are free of functions contained in the PO and all
% activition records are free of functions that $(1)$ access any global
% or static symbols we identify during Katana's {\it Data Traversal} stage and
% $(2)$ do not define any local variables of modified types identified
% during our Type Discovery phase.  Katana uses the {\tt ptrace}
% interface to pause execution and query this state.
%
%JMHO: We repeat this again in the discussion section, so I commented
%it out here and added a brief note saying we discuss it more below
We have not
yet implemented a safety determination mechanism, but we discuss our
approach to it in the next section.

>>>>

For dynamic symbols, Katana traverses the PLT entries of the
executable and compares the symbol name of each entry with
the symbol name that requires relocation. Once a match is found, the
symbol value can be determined. Our current prototype cannot add calls
to previously unused functions in shared libraries, but support for
this will be added using the ``ALTPLT'' technique described in \cite{elf-debugging}.

%JMHO: this got too many negative comments from reviewers, removing
%and clarifying above
% Katana is unable to handle calls to previously
% %SB:
% unused functions present in any shared libraries\footnote{This would
%   require creation of new PLT and GOT entries and either subsequent
%   re-basing of the following segments of the executable, or creation
%   of a new segment to allocate the extra entries. Although ELF
%   rewriting systems like ERESI or Diablo show that such manipulations
%   can be made practical, we chose not to complicate Katana with
%   them yet.}
% , although support for this is planned.

@@

(i) {\it Type Discovery:} We set out to discover all newly created or
modified data types -- those that are primarily user-defined (such as
structures and unions in {\bf C}). Katana traverses the type information
(as identified by the above DWARF tags) from the newly created executable,
 and for each encountered type, it searches for
the corresponding type-name within the {\it replicated} executable
(from before the patch). If so found, the full types (i.e. the number,
type and position of all member variables contained within) are
compared to determine if they are identical. If not identical,
a transformation between the old and new versions is generated, along
with DWARF information identifying the type to insert into the PO as
soon as a variable is found making use of the altered type.
%the parent type identifying the struct/union is inserted into the PO.
 Else, if
the type name itself was not found within the replicated executable, then the
current type was created by the patch, and is added as such to the PO.
%\newline
\newline
(ii) {\it Data Traversal:} The next step is to traverse all variables
defined within the new application, and for each one encountered,
we first determine its 
%SB:
lexical scope.
If the scope is
local, then we ensure that the corresponding function (the one that
defines this variable) does not have an activation frame on  
the program stack while
applying the patch. Else, the variable has been defined as either
global or static. We first check if the replicated executable defines
the same variable. If not, then this variable has been created by the
patch and we need not worry about it and leave the symbol resolution upto
the compiler (as only {\it new} code can use this variable). Otherwise,
we verify whether the variable's type is one of the
modified types identified during {\it type discovery}. If it is, then
we add the variable along with its {\it original} address from the replicated
executable, its {\it new} address from the patch, and type information to the
PO. At the end of this stage, Katana would have identified all newly
created or modified variables from the patch. 
%\newline
\newline
(iii) {\it Patch Application: } Applying a data patch consists of
first tracking down the relevant symbols in program memory. Katana
reads in the PO, and for each data variable encountered, it checks if
the variable is a pointer or not. If it is, then the current validity of the
pointer is verified (by bounds-checking the pointer value to within
heap boundaries).  If the pointer is found to be invalid, no further
action is taken.  If the pointer is valid, then memory for the new
type(s) is allocated, the older structure is copied into the new one
taking into account the difference in structure definition, the old
memory is then freed, and the pointer is modified to point to the new
segment (in case of structures such as lists, trees, since we have the
type specification, we can repeat this process recursively for each
node on the list or tree). Else if the variable is not a pointer, then
Katana modifies all its references in the program text to the updated
memory location from the patch. Katana supports default values in the
sense that if the variable has an initializer in the new version of
the program text, that initializer will be used for any member
variables within structures which or array elements which did not
previously exist in the executing target. Initializers from the new
version are used for all \texttt{const} global variables.
Eventually Katana it will support default values and programmer-written
custom initializers.

{\bf Challenges.} %\hskip 0.1in 
Data patching, as described above, is not an easy task, even with
DWARF type information. C, as well as many other directly compiled
languages, does not have a strong type system, and is not designed to
allow reflection. Structures are relatively straightforward to patch,
because they contain a detailed specification. Unions and arrays, on
the other hand, are generally quite opaque. If substantial changes
have been made to one of the types in a union, Katana cannot do
anything automatically, as there is no automatic way to determine
which unioned type to act on. If an array changes size, Katana will
assume that it is growing or shrinking at the end, and copy old data
accordingly, but will issue a warning that this may not be the desired
behavior. Pointers are of course handled, but with some limitations
currently. Unfortunately for our purposes, memory management is not
part of the ABI. Further, there is no standard way to determine what
block a given pointer is part of. Our current implementation correctly
handles only memory management with \texttt{malloc} and only pointers to the
beginning of blocks. Improvement of this is a major area of future
work. Multiple pointers to the same address are handled by keeping track
of which addresses have been relocated. \texttt{void}* also poses a
problem, as there is generally no way to determine the ``real'' type
of the pointer. The general solution to all of these type problems is
to ask the programmer for routines which perform the
application-specific work. Minimizing programmer work is of foremost
importance, because greater human interaction adds greater possibility
for human error, thus possibly decreasing the reliability of the
patch, but it is not avoidable in all situations.

Hot patching still faces a number of other
challenges, including dealing with multithreaded programs and address
space randomization (which slight changes to the OS loader can help us
overcome).  
There is nothing inherent in our design which will not work with
multithreading, but dealing with it through ptrace takes considerable
work which we have not yet done. More importantly, the deadlock
avoidance work required to ensure that we do at some point pause all
threads, but that we never pause a thread that another thread may be
waiting on in order to reach a safe state is nontrivial. We plan to
address this in the future.
% We consider the program to be in a safe state if all
% activation records are free of functions contained in the PO and all
% activition records are free of functions that $(1)$ access any global
% or static symbols we identify during Katana's {\it Data Traversal} stage and
% $(2)$ do not define any local variables of modified types identified
% during our Type Discovery phase.  Katana uses the {\tt ptrace}
% interface to pause execution and query this state.
%
%JMHO: We repeat this again in the discussion section, so I commented
%it out here and added a brief note saying we discuss it more below
